Wait-free Queues with Polylogarithmic Step Complexity

### 2022-10-04 thesis/thesis.tex
add hyperref
add current changes

\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{showkeys}
\usepackage{amsmath}

@@

We could implement the sequence stored in each node using an array of the queue operations and append some operations to the sequence by doing \tt{k-CAS} operation on the end of the array. To do a propagate step on node $n$ in the tree, we aggregate the operations from node $n$'s both children (that have not already been propagated to $n$) and try to append them into $n$. We call this procedure \sc{Refresh}$(n)$. The main idea is that if we call \sc{Refresh}$(n)$ twice, the operations in $n$'s children before the first \sc{Refresh}$(n)$ are guaranteed to be in $n$.
Because if both of the \sc{Refresh}$(n)$s fail to do \nf{k-CAS} then there is another instance of \nf{Refresh()} in between which has succeeded to do \nf{CAS} and has already appended the operations that the first \nf{Refresh} was trying to append. This mechanism makes us overcome the \nf{CAS} Retry Problem.

@@

The solution for  implementing the orderings in the tree told above is not efficient, because there are big \nf{CAS}es and operations information are copied all the way up to the root.
Instead of storing operations explicitly in the nodes, we can keep track of some statistics of them. This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$). To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the statistics block of the new operations. In each \nf{Refresh()} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently. Furthermore since the operations in a \textsc{Refresh()} step are concurrent we can linearize them  among themselves in any order we wish. Note that if two operations are in read one \textsc{Refresh()} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

@@

\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children. \texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block. Other attributes are assigned values followed by definition.
\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/createblock}
  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/doublerefresh.png}
  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
\end{figure}

@@

 Let $i$ be the value $R_n$, a successful instance of \nf{Refresh()} on node \nf{n} reads from \nf{$n$.head}. $R_n$ does a successful \nf{CAS(null, $b$)} into \nf{n.blocks[$i$]}. Let $p$ be $n$\nf{}.parent. Without loss of generality for the rest of this section assume ${n}$ is the left child of $p$. From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh()} propagating $b$. Let $R_p$ be the first successful \nf{p.Refresh()} that reads some value greater than ${i}$ for \nf{left.head} and contains ${b}$ in its created block in Line \ref{invokeCreateBlock}. Let the index of the block $R_p$ put in $p$\nf{.blocks} be $j$.

@@

Block $b$ is appended to \nf{n.blocks[h]} on Line \ref{cas}. After appending $b$, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh()} by the same process or another process or  maybe an \nf{$n$.parent.Refresh()}. We want to bound how far \nf{b.super} is from the index of $b$'s superblock, which is created by a successful \nf{n.parent.Refresh()} that propagates $b$.

@@

\paragraph{}
In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate()} steps and increase parallelism. So now the problem is to find the result of each Dequeue. From lemma \ref{blockSize} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations. In the next example, operations in a cell are concurrent. \texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively. We will next describe how these values can be computed efficiently.

@@

We store ordering among \texttt{operation}s in the tournament tree constructed by \texttt{node}s. In each \texttt{node} we store pointers to its relatives, an array of \texttt{block}s and an index to the first empty \texttt{block}. Furthermore in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}. There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh()} step. It means after that some bunch of \texttt{block}s in a node have propagated into the parent then the \texttt{counter} increases. Each new \texttt{block} added to a node sets its \texttt{time} regarding \texttt{counter}. This helps us to know which blocks have aggregated together to a block, not precisely though. We also store the index of the aggregated \texttt{block} of a \texttt{block} with \texttt{time} $i$ in \texttt{super[i]}.

@@

\texttt{Append(op)} adds \texttt{op} to the invoking process's leaf's \texttt{ops} and \texttt{blocks}, \texttt{propagate}s it up to the root and if the \texttt{op} is a dequeue returns its order in residing block in the root and the block's index. As we said later \texttt{Propagate()} assuredly aggregates new blocks to a block in the parent by calling \texttt{Refresh()} two times. \texttt{Refresh(n)} creates a block, tries to CAS it into the p\texttt{n}'s \texttt{blocks} and if it was successful updates \texttt{super} and \texttt{counter} in both of \texttt{n}'s children.

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{enq}$\geq$i}.}
\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_{this,b,i}=D_{root,x,y}$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= (parent.left==n)? left: right} \cmt{check if this node is a left or a right child}
\State \tt{superBlock= parent.\Call{BinarySearch}{sum\sub{deq-dir}, i+blocks[b-1].sum\sub{deq}, blocks[b].super, blocks[b].super+1}} \label{computeSuper}

\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-left}} \cmt{consider the enqueues in the previous blocks from the left child} \label{considerPreviousLeft}
\EndIf

\If{\tt{dir {\keywordfont is} right}} \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{enq}-blocks[superBlock-1].sum\sub{enq-right}} \cmt{consider the enqueues in the previous blocks from the right child} \label{considerPreviousRight}
\State \tt{i+= blocks[superBlock].num\sub{deq-left}} \cmt{consider the dequeues from the right child} \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDequeue}{superBlock, i}
\EndIf
\EndFunction{IndexDequeue}

@@

Next, we show that the value of \nf{head} in a node can only be increased. By the termination of a \nf{Refresh()}, \nf{head} has been incremented by the process doing the \nf{Refresh()} or by another process.
\begin{observation} \label{nonDecreasingHead}
For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
\end{observation}
\begin{proof}
  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance()}.
\end{proof}

@@

The value of \nf{n.head} is modified only in Line \ref{incrementHead} of \nf{Advance()}. If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}. \nf{Advance()} is called  in Lines \ref{helpAdvance} and \ref{advance}. For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}. For Line \ref{advance}, Line \ref{cas} was finished  before doing \ref{advance}. Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}. 
\end{proof}

@@

\begin{definition} [Successful Refresh]
  An instance of \nf{Refresh()} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}. If a successful instance of \nf{Refresh()} terminates, we say it is \it{complete.}
\end{definition}

In the next two results we show for every successful \nf{Refresh()}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.

\begin{lemma} \label{lem::trueRefresh}
If $R$ is a successful instance of \nf{$n$.Refresh()}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
\end{lemma}
\begin{proof}

@@

\begin{corollary}\label{lem::prectrueRefresh}
If $R$ is a complete instance \nf{$n$.Refresh()}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}$.
\end{corollary}
\begin{proof}
The left hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$. Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.  Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates. So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$. 
\end{proof}

@@

In the next lemma we show that if two consecutive instances of \nf{Refresh()} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh()} are guaranteed to be in $n$ after the second \nf{Refresh()}.

\begin{lemma} \label{doubleRefresh}
  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh()} on internal node $n$ by process $p$. If neither $R_1$ nor $R_2$ is a successful \nf{Refresh()}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
\end{lemma}
\begin{proof}
Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}. By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ both cannot read the same value $i$. By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.

Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}. As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh()} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}. $R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}. By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}). By Lemma \ref{lem::prectrueRefresh} we have 
$EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$. Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{n.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by the time $R_2$ terminates. To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Observation and Lemma \ref{lem::headInc} we have \ref{lem::blocksOrder} $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n.\nf{left}}^{^{R_2}t}$.

   If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented  more than two times since $_{\ref{readHead}}^{R_1}t$. By Lemma \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[i+1]}$ is non-null. Let $R_3$ be the \nf{Refresh()} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$. $R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}. So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^R_{2}t$. From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh}, \ref{lem::headInc} on $R_3$ the claim holds.
\end{proof}

@@

\begin{corollary} \label{lem::propagateStep}
  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
\end{corollary}
\begin{proof}
  If the first \nf{Refresh()} in line \ref{firstRefresh} returns \nf{true} then by Lemma \ref{lem::prectrueRefresh} the claim holds. If the first \nf{Refresh()} failed and the second \nf{Refresh()} succeeded the claim still holds by Lemma \ref{lem::prectrueRefresh}. Otherwise both failed and the claim is satisfied by Lemma \ref{doubleRefresh}.
\end{proof}

@@

\begin{corollary}\label{lem::append}
For $A=l\nf{.Append(}b\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
\end{corollary}
\begin{proof}
$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate()} on the parent of the leaf where it appended $b$. For every node $n$, $n.$\nf{Propagate()} appends $b$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate()} untill $n$ is \nf{root}.
\end{proof}

@@

First we prove some claims about the size and operations of a block. These lemmas will be used later for correctness and analysis of \nf{GetEnqueue()}.

\begin{lemma}\label{blockSize}
Each block contains at most one operation of each process.
\end{lemma}
\begin{proof}
To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $p$ in block $b$ in node $n$. Without loss of generality $op_1$ is invoked earlier than $op_2$. Process $p$ cannot invoke more than one operations concurrently, so $op_1$ has to be finished before $op_2$ begins. By Corollary \ref{lem::appendExactlyOnce} before $op_2$ calls \nf{Append()}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root. This means there is some block $b^\prime$ before $b$ in $n$ containing $op_1$. The Existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
\end{proof}

\begin{lemma}\label{subBlocksBound}
Each block has at most $p$ direct subblocks.
\end{lemma}
\begin{proof}
The claim follows directly from Lemma \ref{blockSize} and the observation that each block appended to an initial node contains at least one operation, due to the test on Line~\ref{addOP}. We can also see the blocks in the leaves have exactly one operation in the \nf{Enqueue()} and \nf{Dequeue()} routines.
\end{proof}

>>>>

\usepackage[letterpaper, margin=1in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
%\usepackage{showkeys}
\usepackage{hyperref}
\usepackage{amsmath}

@@

We could implement the sequence stored in each node using an array of the queue operations and append some operations to the sequence by doing \tt{k-CAS} operation on the end of the array. To do a propagate step on node $n$ in the tree, we aggregate the operations from node $n$'s both children (that have not already been propagated to $n$) and try to append them into $n$. We call this procedure \sc{Refresh}$(n)$. The main idea is that if we call \sc{Refresh}$(n)$ twice, the operations in $n$'s children before the first \sc{Refresh}$(n)$ are guaranteed to be in $n$.
Because if both of the \sc{Refresh}$(n)$s fail to do \nf{k-CAS} then there is another instance of \nf{Refresh} in between which has succeeded to do \nf{CAS} and has already appended the operations that the first \nf{Refresh} was trying to append. This mechanism makes us overcome the \nf{CAS} Retry Problem.

@@

The solution for  implementing the orderings in the tree told above is not efficient, because there are big \nf{CAS}es and operations information are copied all the way up to the root.
Instead of storing operations explicitly in the nodes, we can keep track of some statistics of them. This allows us to \texttt{CAS} fixed-size objects in each \textsc{Refresh}($n$). To do that, we introduce blocks that only contain the number of operations from the left and the right child in a \texttt{Refresh}() procedure and only propagate the statistics block of the new operations. In each \nf{Refresh} there is at most one operation from each process trying to be propagated, because one operation cannot invoke two operations concurrently. Furthermore since the operations in a \textsc{Refresh} step are concurrent we can linearize them  among themselves in any order we wish. Note that if two operations are in read one \textsc{Refresh} step in a node they are going to be propagated up to the root together. Our choice is to put the operations propagated from the left child before the operations propagated from the right child. In this way if we know the number of operations from the left child and the number of operations from the right child in a block  we have a complete ordering on the operations.

@@

\paragraph{\texttt{CreateBlock}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children. \texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block. Other attributes are assigned values followed by definition.
\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/createblock}
  \caption{\label{fig::createBlock}Snapshot of a CreateBlock}
\end{figure}

\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/doublerefresh.png}
  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock}.}
\end{figure}

@@

 Let $i$ be the value $R_n$, a successful instance of \nf{Refresh} on node \nf{n} reads from \nf{$n$.head}. $R_n$ does a successful \nf{CAS(null, $b$)} into \nf{n.blocks[$i$]}. Let $p$ be $n$\nf{}.parent. Without loss of generality for the rest of this section assume ${n}$ is the left child of $p$. From Lemma \ref{lem::appendExactlyOnce} we know there could be only one \nf{p.Refresh} propagating $b$. Let $R_p$ be the first successful \nf{p.Refresh} that reads some value greater than ${i}$ for \nf{left.head} and contains ${b}$ in its created block in Line \ref{invokeCreateBlock}. Let the index of the block $R_p$ put in $p$\nf{.blocks} be $j$.

@@

Block $b$ is appended to \nf{n.blocks[h]} on Line \ref{cas}. After appending $b$, \nf{b.super} is set on Line \ref{setSuper1} of a call to \nf{Advance} from \nf{$n$.Refresh} by the same process or another process or  maybe an \nf{$n$.parent.Refresh}. We want to bound how far \nf{b.super} is from the index of $b$'s superblock, which is created by a successful \nf{n.parent.Refresh} that propagates $b$.

@@

\paragraph{}
In the Block Tree, we did not store the sequence of operations explicitly but instead stored blocks of concurrent operations to optimize \texttt{Propagate} steps and increase parallelism. So now the problem is to find the result of each Dequeue. From lemma \ref{blockSize} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue operations. In the next example, operations in a cell are concurrent. \texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4, null} respectively. We will next describe how these values can be computed efficiently.

@@

We store ordering among \texttt{operation}s in the tournament tree constructed by \texttt{node}s. In each \texttt{node} we store pointers to its relatives, an array of \texttt{block}s and an index to the first empty \texttt{block}. Furthermore in \texttt{leaf} nodes there is an array of \texttt{operations} where each \texttt{operation} is stored in one cell with the same index in \texttt{blocks}. There is a \texttt{counter} in each \texttt{node} incrementing after a successful \texttt{Refresh} step. It means after that some bunch of \texttt{block}s in a node have propagated into the parent then the \texttt{counter} increases. Each new \texttt{block} added to a node sets its \texttt{time} regarding \texttt{counter}. This helps us to know which blocks have aggregated together to a block, not precisely though. We also store the index of the aggregated \texttt{block} of a \texttt{block} with \texttt{time} $i$ in \texttt{super[i]}.

@@

\texttt{Append(op)} adds \texttt{op} to the invoking process's leaf's \texttt{ops} and \texttt{blocks}, \texttt{propagate}s it up to the root and if the \texttt{op} is a dequeue returns its order in residing block in the root and the block's index. As we said later \texttt{Propagate} assuredly aggregates new blocks to a block in the parent by calling \texttt{Refresh} two times. \texttt{Refresh(n)} creates a block, tries to CAS it into the p\texttt{n}'s \texttt{blocks} and if it was successful updates \texttt{super} and \texttt{counter} in both of \texttt{n}'s children.

@@

\Statex
\Statex $\leadsto$ \textsf{Precondition: \tt{b}th block of the node has propagated up to the root and \tt{blocks[b].num\sub{deq}$\geq$i}.}
\Function{<int, int>}{IndexDequeue}{\sl{int} b, \sl{int} i} \cmt{Returns \tt{<x, y>} if $D_i(this, b)=D_y(root,x)$.}
\If{\tt{this} \bf{is} \tt{root}}
\State\Return \tt{<b, i>} \label{indexBaseCase}
\Else
\State \tt{dir= parent.left==n ? left: right} 
\State \tt{sb= parent.blocks[blocks[b].super].sum\sub{deq-dir} > blocks[b].sum\sub{deq} ?  blocks[b].super: blocks[b].super+1}\label{computeSuper}

\If{\tt{dir {\keywordfont is} left}} \label{computeISuperStart}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-left}} \label{considerPreviousLeft}
\EndIf

\If{\tt{dir {\keywordfont is} right}} \label{considerRight}
\State \tt{i+= blocks[b-1].sum\sub{deq}-parent.blocks[sb-1].sum\sub{deq-right}}  \label{considerPreviousRight}
\State \tt{i+= blocks[sb].num\sub{deq-left}}  \label{considerLeftBeforeRight}
\EndIf \label{computeISuperEnd}
\State \Return\tt{this.parent.}\Call{IndexDequeue}{sb, i}
\EndIf
\EndFunction{IndexDequeue}

@@

Next, we show that the value of \nf{head} in a node can only be increased. By the termination of a \nf{Refresh}, \nf{head} has been incremented by the process doing the \nf{Refresh} or by another process.
\begin{observation} \label{nonDecreasingHead}
For each node $n$,  $n.\nf{head}$ is non-decreasing over time.
\end{observation}
\begin{proof}
  The claim follows trivially from the code since \nf{head} is only changed by incrementing in Line \ref{incrementHead} of \nf{Advance}.
\end{proof}

@@

The value of \nf{n.head} is modified only in Line \ref{incrementHead} of \nf{Advance}. If $n$\nf{.head} is incremented to $h+1$ it is sufficient to show \nf{n.blocks[$h$]}$\neq$\nf{null}. \nf{Advance} is called  in Lines \ref{helpAdvance} and \ref{advance}. For Line \ref{helpAdvance}, $n.\nf{blocks[h]}\neq\nf{null}$ because of the \nf{if} condition in Line \ref{ifHeadnotNull}. For Line \ref{advance}, Line \ref{cas} was finished  before doing \ref{advance}. Whether Line \ref{cas} is successful or not, $n.\nf{blocks[h]}\neq\nf{null}$ after the $n$\nf{.blocks[$h$].CAS}. 
\end{proof}

@@

\begin{definition} [Successful Refresh]
  An instance of \nf{Refresh} is \it{successful} if its \nf{CAS} in Line \ref{cas} returns \nf{true}. If a successful instance of \nf{Refresh} terminates, we say it is \it{complete.}
\end{definition}

In the next two results we show for every successful \nf{Refresh}, all the operations established  in the children before the \nf{Refresh} are in the parent after the \nf{Refresh}'s successful \nf{CAS} at Line \ref{cas}.

\begin{lemma} \label{lem::trueRefresh}
If $R$ is a successful instance of \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq ops(n\nf{.blocks}_{\ref{cas}})$.
\end{lemma}
\begin{proof}

@@

\begin{corollary}\label{lem::prectrueRefresh}
If $R$ is a complete instance \nf{$n$.Refresh}, then we have $EST_{n\nf{.left}}^{t^R} \; \cup \; EST_{n\nf{.right}}^{t^R} \subseteq EST_{n}^{^Rt}$.
\end{corollary}
\begin{proof}
The left hand side is the same as Lemma \ref{lem::trueRefresh}, so it is sufficient to show when $R$ terminates the established blocks in $n$ are a superset of $n$\nf{.blocks}$_{\ref{cas}}$. Line \ref{cas} writes the block \nf{new} in $n$\nf{.blocks[$h$]} where $h$ is value of $n$\nf{.head} read at Line \ref{readHead}.  Because of Lemma \ref{lem::headInc}  we are sure that $n\nf{.head}>h$ when $R$ terminates. So the block \nf{new} appended to $n$ at Line \ref{cas} is established at $^Rt$. 
\end{proof}

@@

In the next lemma we show that if two consecutive instances of \nf{Refresh} by the same process on node $n$ fail, then the blocks established in the children of $n$ before the first \nf{Refresh} are guaranteed to be in $n$ after the second \nf{Refresh}.

\begin{lemma} \label{doubleRefresh}
  Consider two consecutive terminating instances $R_1$, $R_2$ of \nf{Refresh} on internal node $n$ by process $p$. If neither $R_1$ nor $R_2$ is a successful \nf{Refresh}, then we have $EST_{n\nf{.left}}^{t^{R{_1}}} \; \cup \; EST_{n\nf{.right}}^{t^{R{_1}}} \subseteq EST_{n}^{^{R{_2}}t}$.
\end{lemma}
\begin{proof}
Let $R_1$ read $i$ from $n\nf{.head}$ at Line \ref{readHead}. By Lemma \ref{lem::headInc}, $R_1$ and $R_2$ both cannot read the same value $i$. By Observation \ref{nonDecreasingHead}, $R_2$ reads a larger value of $n$\nf{.head} than $R_1$.

Consider the case where $R_1$ reads $i$ and $R_2$ reads $i+1$ from Line \ref{readHead}. As $R_2$'s \nf{CAS} in Line \ref{cas} returns \nf{false}, there is another successful instance $R_2^\prime$ of \nf{$n$.Refresh} that has done a \nf{CAS} successfully into \nf{$n$.blocks[$i+1$]} before $R_2$ tries to \nf{CAS}. $R_2^\prime$ creates its block \nf{new} after reading the value $i+1$ from $n$\nf{.head} (Line \ref{readHead}) and $R_1$ reads the value $i$ from $n$\nf{.head}. By Observation \ref{nonDecreasingHead} we have $^{R_1}t<t^{R_1}_{\ref{readHead}}< t^{R2^\prime}_{\ref{readHead}}$ (see Figure \ref{fig::doubleRefresh}). By Lemma \ref{lem::prectrueRefresh} we have 
$EST^{n.\nf{left}}_{^{R_2^\prime}_{\ref{readHead}}t} \cup EST^{n.\nf{right}}_{^{R_2^\prime}_{\ref{readHead}}t} \subseteq ops(n.\nf{blocks}_{t_{\ref{cas}}^{R_2^\prime}})$. Also by Lemma \ref{lem::headInc} on $R_2$, the value  of \nf{n.head} is more than $i+1$ after $R_2$ terminates, so the block appended by $R_2^\prime$ to \nf{n} is established by the time $R_2$ terminates. To summarize, $^{R_1}t$ is before $R_2^\prime$'s read of \nf{$n$.head} ($t_{\ref{readHead}}^{R_2^\prime}$) and $R_2^\prime$'s successful \nf{CAS} ($t^{R_2^\prime}_{\ref{cas}}$) is before $R_2$'s termination ($t^{R_2}$), so by Observation and Lemma \ref{lem::headInc} we have \ref{lem::blocksOrder} $EST_{n.\nf{left}}^{t^{R_1}} \cup EST_{n.\nf{right}}^{t^{R_1}} \subseteq ops(n.\nf{blocks}_{t^{R_2^\prime}_{\ref{cas}}}) \subseteq EST_{n.\nf{left}}^{^{R_2}t}$.

   If $R_2$ reads some value greater than $i+1$ in Line \ref{readHead} it means $n\nf{.head}$ has been incremented  more than two times since $_{\ref{readHead}}^{R_1}t$. By Lemma \ref{lem::headPosition}, when $n.$\nf{head} is incremented from $i+1$ to $i+2$, $n\nf{.blocks[i+1]}$ is non-null. Let $R_3$ be the \nf{Refresh} on $n$ that has put the block in $n\nf{.blocks[}i+1\nf{]}$. $R_3$ read $n\nf{.head}=i+1$ at Line \ref{readHead} and has put its block in $n\nf{.blocks[}i+1\nf{]}$ before $R_2$'s read of $n.$\nf{head} at Line \ref{readHead}. So we have $t^{R_1}<_{\ref{readHead}}^{R_3}t<_{\ref{cas}}^{R_3}t<t^{R_2}_{\ref{readHead}}<^R_{2}t$. From Observation \ref{lem::blocksOrder}  on the operations before and after $R_3$'s \nf{CAS} and Lemmas \ref{lem::trueRefresh}, \ref{lem::headInc} on $R_3$ the claim holds.
\end{proof}

@@

\begin{corollary} \label{lem::propagateStep}
  $EST_{n\nf{.left}}^{_{\ref{firstRefresh}}t} \cup EST_{n\nf{.right}}^ {_{\ref{firstRefresh}}t} \subseteq EST_{n}^{t_{\ref{secondRefresh}}}$
\end{corollary}
\begin{proof}
  If the first \nf{Refresh} in line \ref{firstRefresh} returns \nf{true} then by Lemma \ref{lem::prectrueRefresh} the claim holds. If the first \nf{Refresh} failed and the second \nf{Refresh} succeeded the claim still holds by Lemma \ref{lem::prectrueRefresh}. Otherwise both failed and the claim is satisfied by Lemma \ref{doubleRefresh}.
\end{proof}

@@

\begin{corollary}\label{lem::append}
For $A=l\nf{.Append(}b\nf{)}$ we have ${ops(b)} \subseteq EST_{n}^{t^A}$ for each node $n$ in the path from $l$ to \nf{root}.
\end{corollary}
\begin{proof}
$A$ adds $b$ to the assigned leaf of the process, establishes it at Line \ref{appendEnd} and then calls \nf{Propagate} on the parent of the leaf where it appended $b$. For every node $n$, $n.$\nf{Propagate} appends $b$ to $n$, establishes it in $n$ by Corollary \ref{lem::propagateStep} and then calls $n.$\nf{parent.Propagate} untill $n$ is \nf{root}.
\end{proof}

@@

First we prove some claims about the size and operations of a block. These lemmas will be used later for the correctness and analysis of \nf{GetEnqueue()}.

\begin{lemma}\label{blockSize}
Each block contains at most one operation of each process, and therefore at most $p$ operations in total.
\end{lemma}
\begin{proof}
To derive a contradiction, assume there are two operations $op_1$ and $op_2$ of process $p$ in block $b$ in node $n$. Without loss of generality $op_1$ is invoked earlier than $op_2$. Process $p$ cannot invoke more than one operation concurrently, so $op_1$ has to be finished before $op_2$ begins. By Corollary \ref{lem::appendExactlyOnce} before $op_2$ calls \nf{Append}, $op_1$ exists in every node of the tree on the path from $p$'s leaf to the root. Since $b$ contains $op_2$, it must be created after $op_2$ is invoked. This means there is some block $b^\prime$ before $b$ in $n$ containing $op_1$. The existence of $op_1$ in $b$ and $b^\prime$ contradicts Lemma \ref{lem::noDuplicates}.
\end{proof}

\begin{lemma}\label{subBlocksBound}
Each block has at most $p$ direct subblocks.
\end{lemma}
\begin{proof}
The claim follows directly from Lemma \ref{blockSize} and the observation that each block appended to an internal node contains at least one operation, due to the test on Line~\ref{addOP}. We can also see the blocks in the leaves have exactly one operation in the \nf{Enqueue()} and \nf{Dequeue()} routines.
\end{proof}
